Assessing the Impact of MDE and Code Generation on the Sustainability of Scientific Computing Software: A Research Proposal

### 2019-09-23 ResearchProposal.tex
Updates to research proposal

\section{Introduction}

@@

This section is decomposed into 4 subsections -- one for each of the different
approaches for assessing the impact of MDE on the sustainability of SCS.  The
goal of the first section is to establish a ``baseline'' on the current state of
the practice for SCS.  This is done by identifying domains within SCS where
significant software development work has been done and then systematically
assessing the quality of the existing work.  The next section looks specifically
at the impact of MDE on SCS through a series of case studies.  An ``ideal'' MDE
process is proposed and then followed with different stakeholders and
qualitative data is collected.  The last two sub-sections discuss assessing the
impact on SCS of assurance cases and computational variability testing,
respectively.

@@

A survey of existing SCS software will identify patterns, find opportunities for
improvement and identify candidate program families for future implementation.
The survey will require developing terminology and a classification system to
answer such questions as: What are the current best practises?  What software
packages are reused and why these packages?  What do practitioners look for in
the tools that support their work?

To achieve objective S2, a measurement template, possibly using pairwise
comparisons of quality metrics between software packages, will be developed for
assessing the quality of the documentation, code, test cases and development
processes for existing SCS program families.  The quality measurement template
will be used to assess the quality of approximately 30 different SCS families.
About half of the families will be model oriented and the other half tool
oriented.  Additional details on best practices will be collected by students
interviewing code owners.  Based on preliminary discussions, Dr.\ Jeffrey Carver
from the University of Alabama and some of his colleagues may also participate
in collecting interview data. Once all the surveys and interviews are
completed, a meta-analysis will draw conclusions for each domain and between
domains.  Knowing the state of practice for family development will highlight
which domains are using best practices that should be emulated in LSS.

The plan here is to return to the task of measuring/assessing the state of
software development practice in several scientific computing domains.  We will
update the work that was done previously for domains such as Geographic
Information Systems~\cite{SmithEtAl2018_arXivGIS}, Mesh
Generators~\cite{SmithEtAl2016}, Seismology software~\cite{SmithEtAl2018}, and
Statistical software for psychology~\cite{SmithEtAl2018_StatSoft}.  We could
return to these domains and/or introduce new domains.  Potential domains/collaborators
are listed in the subsections.

MEng students will be asked to assess the state of practice in each domain.
Last time each student measured two domains, so we would start off with this
model again.  If the scope becomes large enough, we might switch to one domain
per student.

In the previous project, we measured 30 software projects for each domain.  With
the increased scrutiny required in this re-boot, we won't likely be able to
measure the details on this high a number of software projects.  We may still
start with 30 software examples in each domain, but then use some criteria to
create a shorter sub-list for detailed measurement.  The details of how to
proceed here would be determined and documented in the measurement protocol.

In the previous state of practice assessing exercise, a series of questions and
simple metrics was devised to measure the quality of the documentation and
adherence to best software development practices.  The new project will
critically assess the previous set of questions and revise as necessary.  In
addition, the following data will be collected/developed for each domain:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain.  What services does each member of the domain provide?  Ideally this
  information would be summarized in a commonality analysis document.  The
  document will summarize the domain software via its commonalities,
  variabilities and parameters of variation.
\item Usability testing of each software in the domain.  Specific metrics for
  assessing usability still need to be researched, but measures that involve the
  time to complete a task and the users overall impression would be considered.
\item Collect empirical software engineering related data, such as the number of
  files, number of lines of code, cyclomatic complexity, number of open issues,
  etc.  As a starting point for tool support, HubListener
  (\url{https://github.com/pjmc-oliveira/HubListener}) could be used.
\item Collect at least one empirical measure of the quality of the documentation
  - the number of lines of documentation.  [Has anyone previously looked at this
  metric?]
\end{itemize}

The above data will be combined to rank the software in the domain.  The
specifics are yet to be determined, but in the past the Analytic Hierarchy
Process proved helpful in this.

@@

GitHub will be used to coordinate the work of the large team of people that will
be involved in this project.  In addition to the project record left on GitHub,
the final data will be exported to Mendeley.

Determine criteria for what makes a domain a good fit.

Write up experimental procedure as a publication - on arxiv, but hopefully
somewhere else too.

Start from previous work and modify.  Brainstorm how to do measures.

Schedule meeting.

Potential collaborators/domains are listed in the following subsections.

Rule 4: Keep knowledge up to date and findable \citep{ShollerEtAl2019}.

@@

Medical imaging and analysis software (Dr.\ Michael Noseworthy)?

@@

Climate modelling software (Dr.\ Zoe Li or Dr.\ Xander Wang)?

@@

Extraction of 3D geometries from medical images (Dr.\ Zahra Motamed)?  Lattice
Boltzmann Solver?

@@

Finite element solvers and/or mesh generators (Dr.\ Dieter Stolle)?  Potential
software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
and use them to do something non-trivial.

@@

Statistics software for psychology (Dr.\ Karin Humphreys)?

@@

Li Xi (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?

\subsubsection{Geographic Information Systems}

Geographic Information Systems (Jason Brodeur and/or Patrick DeLuca)?

@@

Solidification and casting software (Dr.\ Andr\'e Phillion)?

@@

Quantum chemistry software (Dr.\ Paul Ayers)?

\subsection{Impact of MDE on SCS} \label{SecImpactMDE}

roadmap

Research Question: What tools and methods from MDE are perceived by end user developers to
  improve their productivity and the sustainability of their software.

\subsubsection{Fitting Drasil into the Scientific Software Development Process} \label{Sec}

Scientists generally use an approximation of agile methods for their software
development.  In addition, documentation is generally not emphasized.  How
should Drasil be fit into the scientific software development process?  The
Drasil process should encourage an early intensity of thought and effort, but at
the same time, it shouldn't require waterfall development.  Drasil supports
frequent chance and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
Drasil and then borrow from existing knowledge and create new knowledge?  This
project could start with something like the
\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
{Projectile} example currently in Drasil.

Drasil process should move importance of requirements and domain analysis
earlier in the process.  The quality of the resulting software depends on the
quality of the requirements.  As is known from historical data, the greatest
return on investment is at the requirements state (Boehm papers).
Unfortunately, developers don't like requirements.  Quite possibly they don't
like the intensity of thought that is necessary to get the requirements right.
Inspection techniques, such as task based inspection (Kelly et al) hold promise
for getting the scientist to think deeply at an earlier stage in the development
process.

\subsubsection{etc.}

@@

Empirical Measurement of Documentation: Considerable effort has been invested
into developing metrics, like cyclomatic complexity, to attempt to quantify code
quality, but no measures (as far as I know) exist for quantifying the quality of
documentation.  (Khedri does have some work with Bhahati (sp?) Sanga that
quantifies the connectivity of the documentation via the traceability graph.)
Likely the reason that effort has not been invested to quantifying the quality
of documentation is that is not represented formally.  With Drasil much of the
documentation is formal.  This should facilitate automatic measures of
documentation quality, or at least metrics that might be correlated with
quality.

Measuring the Impact of Drasil on Developing Scientific Software: This project
is challenging to design, but would be HUGE if we were successful.  We assume
that Drasil will improve the quality of scientific software and the efficiency
of its production, but data to back this up would go a long way to convincing
the scientific community.  Experiments where a given project is developed in
parallel by two independent teams come to mind, but this would be expensive, and
difficult to control.  Some additional thought, and resources, are necessary
here.  If we could do a project where the scientist, or a student under their
supervision, is a true partner in the development, we might get more convincing
anecdotal evidence.

>>>>

\section{Introduction} \label{SecIntroduction}

@@

This section is decomposed into 4 subsections -- one for each of the different
approaches for assessing the impact of MDE on the sustainability of SCS, as
outlined in the Introduction (Section~\ref{SecIntroduction}).  The goal of the
first subsection is to establish a ``baseline'' on the current state of the
practice for SCS.  This is done by identifying domains within SCS and then
systematically assessing the quality of the work in those domains.  The second
subsection looks specifically at the impact of MDE on SCS through a series of
case studies.  An ``ideal'' MDE process is proposed and then followed with
different stakeholders and qualitative data is collected.  The last two
sub-sections discuss assessing the impact on SCS of assurance cases and
computational variability testing, respectively.

@@

To understand the current quality of SCS and to assist in developing methods for
measuring qualities, we will identify several domains within SCS and endeavour
to summarize the current state of the practice within those domains.  This
``state of the practice'' exercise will build off of prior work on
measuring/assessing the state of software development practice in several SCS
domains.  We will update the work that was done previously for domains such as
Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators
\citep{SmithEtAl2016}, Seismology software \citep{SmithEtAl2018}, and
Statistical software for psychology \citep{SmithEtAl2018_StatSoft}.  We may
return to these domains and/or introduce new domains.  Potential
domains/collaborators are listed in the subsections below.

MEng students will be asked to assess the state of practice in each domain.
During the last state of the practice exercise each student measured two
domains, but with the higher demands from the new measurement protocol, we will
likely switch to one domain per student.

In the previous project, we measured 30 software projects for each domain.  With
the increased scrutiny required in this re-boot, we will not likely be able to
measure the details on this high a number of software projects.  We may still
start with 30 software examples in each domain, but then use some criteria to
create a shorter sub-list for detailed measurement.  The details of how to
proceed here would be determined and documented in the measurement protocol.

In the previous state of the practice assessing exercise
\citep{SmithEtAl2018_arXivGIS}, a series of questions and simple metrics was
devised to measure the quality of the documentation and adherence to best
software development practices.  The previous set of questions is provided
\href{https://github.com/adamlazz/DomainX/blob/master/TemplateToGradeSCSoft.pdf}
{here}. The new project will critically assess the previous set of questions and
revise as necessary.  In addition, the following data will also be considered
for collection for each domain:

\begin{itemize}
\item Characterization of the functionality provided by the software in the
  domain.  What services does each member of the domain provide?  Ideally this
  information would be summarized in a commonality analysis document.  The
  document will summarize the domain software via its commonalities,
  variabilities and parameters of variation.
\item Usability testing of each software in the domain.  Specific metrics for
  assessing usability still need to be researched, but measures that involve the
  time to complete a task and the users overall impression will be considered.
\item Collect empirical software engineering related data, such as the number of
  files, number of lines of code, cyclomatic complexity, number of open issues,
  etc.  As a starting point for tool support, HubListener
  (\url{https://github.com/pjmc-oliveira/HubListener}) could be used.
\item Collect at least one empirical measure of the quality of the documentation
  - the number of lines of documentation.  We should investigate to see if this
  is a metric that has previously been collected.  So many of the existing
  metrics are completely focused on the code, or possibly on issue creation and
  management; it seems that little work has been done to assess the
  documentation quality.
\item One of the guidelines for successfully on-boarding new developers is to
  ``Keep knowledge up to date and findable'' \citep{ShollerEtAl2019}.  The
  measurements selected should be able to assist with judging whether a project
  succeeds on this front.
\end{itemize}

The above data will be combined to rank the software in the domain.  The
specifics are yet to be determined, but in the past the Analytic Hierarchy
Process (AHP) \citep{Saaty1980} proved helpful in this.  In the past the AHP
results were calculated by a grade out of 10 that each student evaluator gave to
each quality for each measured product.  A better approach may be to have a
group of people familiar with the software do a pairwise comparison between each
product to come up with a number.  The number would be based on the data
collected, but not in a numerical way.  The data collected would sever the same
role as a decision support system.  It would provided insight, but human beings
would do the final pairwise comparisons.

@@

GitHub will be used to coordinate the work of the large team of people that will
be involved in this project.  In addition to the project record left on GitHub,
the final data will be exported to Mendeley (as was done for the previous
quality measurement exercise, for instance the grades for GIS software are
summarized at \href{https://data.mendeley.com/datasets/6kprpvv7r7/1}
{https://data.mendeley.com/datasets/6kprpvv7r7/1}.)

As a next step for this work, the procedure for measuring software qualities in
SCS will be brainstormed, discussed and written up.  The procedure will be
applied to each domain with SCS that is assessed.  The write-up on the procedure
will be published.  Ideally at a relevant conference, but if that is not a
feasible option, it will be hosted on the arxiv.

Potential domains for future measurement are listed below.  Determination of
these domains are good candidates will require evaluation of the following
criteria:

\begin{itemize}
\item Availability of a domain expert that is willing to assist, and who has at
  least a few graduate students that can assist with the measurement tasks.
\item A domain where end-user developed software is common
\item A domain where open-source software is available.  Commercial products are
  not as effective for measurement, since we cannot generally gain access to the
  source code, developer documentation, or issue tracking system.
\item A domain where scientific computing is used.  The selected software should
  fit in the larger domain of SCS.
\end{itemize}

The data for each of the different subdomains within SCS will be standardized to
facilitate a meta-analysis that combines all of the data.

@@

Medical imaging and analysis software.  This work could build off of the
previous project on this topic completed by Vasudha Kapil.  (Vasudha's report is
located in the \texttt{mmsc} svn repo) .  % (Dr.\ Michael Noseworthy)?

@@

Climate modelling software.  This could be software used for multi-model
ensemble aggregating \citep{SamoulyEtAl2018} or possibly with software that uses
physical models to predict climate change.
% (Dr.\ Zoe Li or Dr.\ Xander Wang)?

@@

Potential sub-domains with computational medicine could include: extraction of
3D geometries from medical images, or the use of Lattice Boltzmann
Solvers.  % (Dr.\ Zahra Motamed)?

@@

Finite element solvers and/or mesh generators.  Potential
software includes FeNiCS, MOOSE, FreeFEM++, Dolphin, etc.  Install the packages
and use them to do something non-trivial.  %(Dr.\ Dieter Stolle)?

@@

Statistics software for psychology.  A project in this domain would allow us to
update \citet{SmithEtAl2018_StatSoft}. %(Dr.\ Karin Humphreys?)

@@

Software is used for many purposes in chemical engineering.  We would want to
find an application where non-commercial software is available.
% Li Xi? (\url{https://www.eng.mcmaster.ca/chemeng/people/faculty/li-xi})?

\subsubsection{Geographic Information Systems (GIS)}

We studied GIS software in the past \citep{SmithEtAl2018_arXivGIS}, so this
would be an interesting area to update.  The key for the updated effort will be
to identify a domain expert available for collaboration. %(Jason Brodeur and/or Patrick DeLuca)?

@@

Solidification and casting software and other software used for physical
metallurgy.  An example of software written for analyzing the solidification of
a molten alloy is given in \citet{SmithEtAl2019_arXiv}.  % (Dr.\ Andr\'e Phillion)?

@@

Quantum chemistry software.  One example is software for searching for
conformers, like
\href{https://github.com/PeaWagon/Kaplan}{Kaplan}.  %(Dr.\ Paul Ayers)?

\subsection{Impact of MDE on the Sustainability of SCS} \label{SecImpactMDE}

Once we have details on the current state of the practice in SCS, we can begin
to apply MDE to SCS projects and assess the ramifications of this change.  This
section starts by laying out the details on how MDE can be applied to SCS
(Section~\ref{SecMDEProcess}).  We plan to start with an ideal process and then
approximate it through different experiments to measure and assess the impact of
MDE.  The second subsection (Section~\ref{SecQOfMDEArtifacts}) outlines how the
quality of MDE artifacts can be assessed.  These measurements will be used in
the various experiments to assess the impact of MDE.  The third and fourth
subsections describe two experiments for collecting data: i) conducting case
studies with real SCS projects (Section~\ref{SecCaseStudy}); and, ii) performing
an experiment where multiple developers create a small SCS project in one day
(Section~\ref{SecSmallPhysExample}).

\subsubsection{Fitting MDE into the Scientific Software Development
  Process} \label{SecMDEProcess}

Many researchers have stated that a document driven process is not used by, nor
suitable for, scientific software.  These researchers argue that scientific
developers naturally use an agile philosophy~\cite{AckroydEtAl2008,
  CarverEtAl2007, EasterbrookAndJohns2009, Segal2005}, or an amethododical
process~\cite{Kelly2013}, or a knowledge acquisition driven
process~\cite{Kelly2015}.  However, as discussed in the introduction to this
proposal (Section~\ref{SecIntroduction}, documentation provides many benefits.
An MDE approach holds promise to provide these benefits, while removing the
drudgery from the process of documentation and facilitating inevitable changes
in the documentation and the software.

The general idea that MDE will be helpful is appealing, but to study the impact
of MDE, we need to know how it will actually be used.  How exactly would MDE be
fit into the scientific software development process?  The new process should
encourage an early intensity of thought and effort, but at the same time, it
should not impose waterfall development, since waterfall is an unrealistic
expectation for a problem that is not yet well understood.  MDE supports
frequent change and incremental development, but where should developers start
and how should the work proceed?  How does one create an almost empty project in
MDE and then borrow from existing knowledge and create new knowledge?  To create
a specific MDE process for SCS, the Drasil tools will be utilized.  A good
starting project could be
\href{https://jacquescarette.github.io/Drasil/examples/Projectile/srs/Projectile_SRS.html}
{Projectile}, which models simple projectile motion.

Compared to existing SCS development techniques, the new Drasil process should
emphasize requirements and domain analysis earlier in the process.  The quality
of the resulting software depends on the quality of the requirements.  As
Boehm's data shows~\cite{Boehm1981}, early identification of errors implies a
significant return on investment, since the cost of fixing errors increases
dramatically for later development stages.  An SRS is invaluable for
verification, since the quality of software cannot be assessed without a
standard against which to judge it. Better design decisions are facilitated by
the information captured in an SRS.  Moreover, the SRS aids the software
lifecycle by facilitating incremental development.  This last advantage is
important for SCS software, since changes frequently occur as developers explore
the problem domain.  Unfortunately, developers do not generally write
requirements.  Quite possibly they do not like the intensity of thought that is
necessary to get the requirements right.  Inspection techniques, such as task
based inspection \citep{KellyAndShepard2000, KellyAndShepard2004} hold promise
for getting the scientist to think deeply at an earlier stage in the development
process.

Since Drasil is not the only tool that can be used for MDE, we will also
consider what an MDE approach would look like using other tools.  As a starting
point, we will look at \href{https://www.eclipse.org/epsilon/}{Epsilon}.
Epsilon is a family of languages and tools for code generation, model-to-model
transformation, model validation, comparison, migration and refactoring that
work out of the box with EMF, UML, Simulink, XML and other types of models.

\subsubsection{Measuring the Qualities of MDE Developed
  Artifacts} \label{SecQOfMDEArtifacts}

Some of the quality measures developed in Section~\ref{SecStateOfThePractice}
for the ``state of the practice'' exercise will also be used to measure the
quality of the generated documentation and code.
Section~\ref{SecStateOfThePractice} mentions using code quality metrics, like
cyclomatic complexity, to attempt to quantify code quality.  Given the
importance of documentation, it seems reasonable to also assess its quality.
Measuring documentation quality is not something that is often promoted.
(\citet{Sanga2003} does discuss quantifying the connectivity within an SRS
document by analyzing the traceability graph.)  Part of the reason for the lack
of effort on quantifying the quality of documentation is that is not represented
formally.  With Drasil much of the documentation is formal.  This should
facilitate automatic measures of documentation quality, or at least metrics that
might be correlated with quality.

\subsubsection{Case Study} \label{SecCaseStudy}

For assessing the impact of MDE, experiments where a given project is developed
in parallel by two independent teams are infeasible due to expense and
challenges for having a truly controlled experiment.  Instead, we will focus on
the more practical, but still very useful, case study approach.  Guidelines for
conducting and reporting case study research in software engineering are given
in \citet{RunesonAndHost2009}.  As explained there, a case study should be built
around a research question.  The research question should be specific and
measurable.  For the current case study exercise, we propose the question:
``What tools and methods from MDE are perceived by end user developers to
improve their productivity and the sustainability of their software in the long
term?''

We will identify multiple case studies, with potential projects coming from the
domains listed for the ``state of the practice'' exercise
(Section~\ref{SecStateOfThePractice}), like computational medicine, medical
imaging or quantum chemistry software.  We will first identify the partners and
look for feasible projects that can be developed using an MDE approach.  We will
use the ideal process identified in Section~\ref{SecMDEProcess}.  Compare to Some additional thought, and resources, are necessary
here.  If we could do a project where the scientist, or a student under their
supervision, is a true partner in the development, we might get more convincing
anecdotal evidence.  Pre and post interviews.  Measurement of quality following the
techniques developed in the state of the practice exercise.

List of projects above could also be used for case studies.

Observe - how answer questions when problems come up.  Should have so
pre and post interviews.

\subsubsection{Development of Small Physics
  Examples} \label{SecSmallPhysExample}

describe a different experiment.  conference multiple develop physics code.
Like projectile.  Survey before and after.  Quantify success of efforts and time required.

Sketching out an Assurance Case in advance of developing the software could be
just what is needed to manufacture the up-front intensity that is needed for
successful projects.

@@
